---
title: "Health Insurance Market Place"
output: pdf_document
---

## Objective
Health Insurance Marketplace - dataset from Kaggle
Explore health and dental plans data in the US Health Insurance Marketplace
Do exploration data analysis on the following:- 
1. How do plan rates and benefits vary across states?
2. How do plan benefits relate to plan rates?
3. How do plan rates vary by age?
4. How do plans vary across insurance network providers?
Data Description
This data was originally prepared and released by the Centers for Medicare 
& Medicaid Services (CMS).
 

# Library used
```{r echo = TRUE}
library(plyr)
library(dplyr)
library(ggplot2)
library(maps)
library(reshape2)
```

## Getting,reading and storing data 


```{r echo = TRUE}
setwd("C:/Program Files/RStudio/kaggle/health-insurance-marketplace")
#dataBenefits <- read.csv("BenefitsCostSharing.csv", header = TRUE, stringsAsFactors = FALSE)
dataBusinessRules <- read.csv("BusinessRules.csv", header = TRUE, stringsAsFactors = FALSE)
dataNetwork <- read.csv("Network.csv", header = TRUE, stringsAsFactors = FALSE)
dataPlanAttrib <- read.csv("PlanAttributes.csv", header = TRUE, stringsAsFactors = FALSE)
#dataRate <- read.csv("Rate.csv", header = TRUE, stringsAsFactors = FALSE)
dataSerViceArea <- read.csv("ServiceArea.csv", header = TRUE, stringsAsFactors = FALSE)
#save(dataBenefits, file = "dataBenefits.RData")
#save(dataRate, file = "dataRate.RData")
load("dataBenefits.RData")
load("dataRate.RData")

```

# Exploratory analysis to determine the plans offered across various US states

```{r echo = FALSE}
# create a subset of 2014 to 2016 dataset
planRateBenefit <- subset(dataRate, (dataRate$IndividualRate < "9999" & Age != "Family Option"),
                          select = c(BusinessYear, StateCode, IssuerId,
                                               PlanId, Age, IndividualRate, 
                                               IndividualTobaccoRate))
ByStateCarrier <- aggregate(planRateBenefit[, 2], by = list(StateCode = planRateBenefit$StateCode, 
                                          IssuerId =  planRateBenefit$IssuerId,
                                          PlanId = planRateBenefit$PlanId), length)
```


```{r echo = FALSE}
# Graph - State vs insurance plans offered across various US states
ggplot(data = ByStateCarrier, aes(x = StateCode)) +
  geom_bar(position = "stack") + coord_flip() + xlab("State Code") + ylab("Plans Offered") +
  ggtitle("Health & Dental Plans across various states\nin US Insurance Market Place")

```

# To analyse plan rates across various US states

```{r echo = FALSE}
# create a subset of 2014 to 2016 dataset
VariousRatesPerState <- subset(dataRate, (dataRate$Couple != "NA" |
                        PrimarySubscriberAndOneDependent != "NA" |
                        PrimarySubscriberAndTwoDependents != "NA" |
                        PrimarySubscriberAndThreeOrMoreDependents != "NA" |
                        CoupleAndOneDependent != "NA" |
                        CoupleAndTwoDependents != "NA" |
                        CoupleAndThreeOrMoreDependents != "NA"), 
                        select = c(StateCode, IndividualRate, Couple, 
                                   PrimarySubscriberAndOneDependent,
                                   PrimarySubscriberAndTwoDependents,
                                   PrimarySubscriberAndThreeOrMoreDependents,
                                   CoupleAndOneDependent, CoupleAndTwoDependents,
                                   CoupleAndThreeOrMoreDependents))

TotalRatePerState <- aggregate(VariousRatesPerState[, 1], 
                               by = list(StateCode = VariousRatesPerState$StateCode,
                                         IndRate = VariousRatesPerState$IndividualRate,
                                         Couple = VariousRatesPerState$Couple,
                                         PriSub1Dep = VariousRatesPerState$PrimarySubscriberAndOneDependent,
                                         PriSub2Dep = VariousRatesPerState$PrimarySubscriberAndTwoDependents,
                                         PriSub3Dep = VariousRatesPerState$PrimarySubscriberAndThreeOrMoreDependents,
                                         Cou1Dep = VariousRatesPerState$CoupleAndOneDependent,
                                         Cou2Dep = VariousRatesPerState$CoupleAndTwoDependents,
                                         Cou3Dep = VariousRatesPerState$CoupleAndThreeOrMoreDependents), length)

# Melt the data frame as one 
dfmelt <- melt(TotalRatePerState[,c('StateCode','IndRate','Couple','PriSub1Dep', 'PriSub2Dep',
                                 'PriSub3Dep', 'Cou1Dep', 'Cou2Dep',
                                 'Cou3Dep')], id.vars = 1)
# Plot various rate plans across various US states
# except Individual Tobacco rates as it is NA througout the dataset
ggplot(dfmelt,aes(x = StateCode,y = value/1000)) + 
  geom_bar(aes(fill = variable),position = "stack", stat = "identity") + coord_flip() +
  ylab("Plan rates") + xlab("State Code") +
  ggtitle("Plot of various Health insurance Plan rates\nAcross Various US States") +
  scale_fill_discrete(name = "Rate Plans")

```

# Analyse Plan rate vs age  

```{r echo = FALSE}
# create a subset of 2014 to 2016 dataset
PlanRatevsAge <- subset(dataRate, (dataRate$IndividualRate < "9999" &  StateCode != "NA" |
                 Age != "NA" | IndividualRate != "NA"), 
                 select = c(StateCode, Age, IndividualRate)) 
# Filter data for age group less than equal to 22
TotRatevsAge1 <- filter(PlanRatevsAge, (Age <= "22"))
TotRatevsAge1 <- subset(TotRatevsAge1, !duplicated(TotRatevsAge1[, 3]))
# row 81 producing wrong data so it is removed from the dataset
TotRatevsAge1 <- TotRatevsAge1[-(81),]
# Filter data for age group between 23 and 40
TotRatevsAge2 <- filter(PlanRatevsAge, (Age >= "23" & Age <="40"))
TotRatevsAge2 <- subset(TotRatevsAge2, !duplicated(TotRatevsAge2[, 3]))
# Filter data for age group between 41 and 59
TotRatevsAge3 <- filter(PlanRatevsAge, (Age >= "41" & Age <="59"))
TotRatevsAge3 <- subset(TotRatevsAge3, !duplicated(TotRatevsAge3[, 3]))
# Filter data for Family Option age group
TotRatevsAge4 <- filter(PlanRatevsAge, Age == "Family Option")
TotRatevsAge4 <- subset(TotRatevsAge4, !duplicated(TotRatevsAge4[, 3]))
# Filter data for age group between 60 and 64 
TotRatevsAge5 <- filter(PlanRatevsAge, (Age >= "60" & Age <= "64"))
TotRatevsAge5 <- subset(TotRatevsAge5, !duplicated(TotRatevsAge5[, 3]))
# Filter data for age group 65 and over
TotRatevsAge6 <- filter(PlanRatevsAge, Age == "65 and over")
TotRatevsAge6 <- subset(TotRatevsAge6, !duplicated(TotRatevsAge6[, 3]))
# row 37 producing wrong data so it is removed from the dataset
TotRatevsAge6 <- TotRatevsAge6[-(37),]
```

```{r echo = FALSE}
# Plot 1 - age group 0 to 22
ggplot(TotRatevsAge1, aes(x = StateCode, y = IndividualRate, fill = Age)) +
  geom_bar(position = "dodge", stat = "identity") + 
  ylab("Individual Rates") + xlab("State Code") +
  ggtitle("Plan Rate vs Age across various US States\nfrom 2014 to 2016") +
  theme(text = element_text(size=8))
```

```{r echo = FALSE}
# Plot 2 - age group 23 to 40
ggplot(TotRatevsAge2, aes(x = StateCode, y = IndividualRate/1000, fill = Age)) +
  geom_bar(position = "stack", stat = "identity") +
  ylab("Individual Rates") + xlab("State Code") +
  ggtitle("Plan Rate vs Age across various US States\nfrom 2014 to 2016") +
  theme(text = element_text(size=8))
```

```{r echo = FALSE}
# Plot 3 - age group 41 to 59  
ggplot(TotRatevsAge3, aes(x = StateCode, y = IndividualRate/1000, fill = Age)) +
  geom_bar(position = "stack", stat = "identity") + 
  ylab("Individual Rates") + xlab("State Code") +
  ggtitle("Plan Rate vs Age across various US States\nfrom 2014 to 2016") +
  theme(text = element_text(size=8))  
```

```{r echo = FALSE}
# Plot 4 - age group(Familiy Option)  
ggplot(TotRatevsAge4, aes(x = StateCode, y = IndividualRate, fill = Age)) +
  geom_bar(position = "stack", stat = "identity") +
  ylab("Individual Rates") + xlab("State Code") +
  ggtitle("Plan Rate vs Age across various US States\nfrom 2014 to 2016") +
  theme(text = element_text(size=8))

```

```{r echo = FALSE}
# Plot 5 - age group 60 to 64
ggplot(TotRatevsAge5, aes(x = StateCode, y = IndividualRate/1000, fill = Age)) +
  geom_bar(position = "stack", stat = "identity") +
  ylab("Individual Rates") + xlab("State Code") +
  ggtitle("Plan Rate vs Age across various US States\nfrom 2014 to 2016") +
  theme(text = element_text(size=8))
```

```{r echo = FALSE}
# Plot 6 - age group(65 and over)
ggplot(TotRatevsAge6, aes(x = StateCode, y = IndividualRate/1000, fill = Age)) +
  geom_bar(position = position_stack(), stat = "identity") +
  ylab("Individual Rates") + xlab("State Code") +
  ggtitle("Plan Rate vs Age across various US States\nfrom 2014 to 2016") +
  theme(text = element_text(size=8))
```

# Categorise health plan sold to individuals and small group
# market into coverage tiers based on Issuer Actuarial value
# Bronze - 60%, Silver - 70%, Gold - 80%, Platinum - 90% 
# Catastrophic - below 60% sold to individuals 
```{r echo = FALSE}

IssActuCov <- subset(dataPlanAttrib, (dataPlanAttrib$IssuerActuarialValue != ""),
              select = c("StateCode", "IssuerId", "ServiceAreaId", 
                         "IssuerActuarialValue", "MarketCoverage"))
IssActuCov$IssuerActuarialValue <- gsub("%", "", IssActuCov$IssuerActuarialValue)
IssActuCov$IssuerActuarialValue <- as.integer(IssActuCov$IssuerActuarialValue)
# Bronze Coverage Tier 
BronzeTier <- filter(IssActuCov, (IssuerActuarialValue >= 60 & 
                                    IssuerActuarialValue < 70))
BronzeTier <- BronzeTier %>%
  group_by(StateCode) %>% 
  summarise(BronzeTierCov = mean(IssuerActuarialValue))
# Silver Coverage Tier
SilverTier <- filter(IssActuCov, (IssuerActuarialValue >= 70 & 
                                        IssuerActuarialValue < 80))
SilverTier <- SilverTier %>%
  group_by(StateCode) %>% 
  summarise(SilverTierCov = mean(IssuerActuarialValue))
# Gold Coverage Tier
GoldTier <- filter(IssActuCov, (IssuerActuarialValue >= 80 & 
                                        IssuerActuarialValue < 90))
GoldTier <- GoldTier %>%
  group_by(StateCode) %>% 
  summarise(GoldTierCov = mean(IssuerActuarialValue))
# Platinum Coverage tier
PlatinumTier <- filter(IssActuCov, (IssuerActuarialValue >= 90 &
                                      IssuerActuarialValue <= 100))
PlatinumTier <- PlatinumTier %>%
  group_by(StateCode) %>% 
  summarise(PlatinumTierCov = mean(IssuerActuarialValue))
# Catastrophic Tier falls below 60% Actuarial Value
CatasTier <- filter(IssActuCov, (IssuerActuarialValue < 60))
CatasTier <- CatasTier %>%
  group_by(StateCode) %>% 
  summarise(CatasTierCov = mean(IssuerActuarialValue))
# join all 5 coverage tiers 
CombinedCovTier <- join_all(list(BronzeTier, SilverTier, GoldTier, PlatinumTier, CatasTier), 
                            by = "StateCode")

```

```{r echo = FALSE}
# load US map
MapInsData <- CombinedCovTier
#MapInsData <- group_by(MapInsData, StateCode) %>% 
 # summarise(MedIndRate = median(IndividualRate))
us_state_map <- map_data("state")
MapInsData$region <- tolower(state.name[match(MapInsData$StateCode,state.abb)])
statename <- group_by(us_state_map, region) %>%
  summarise(long = mean(long), lat = mean(lat))
mapdata <- left_join(MapInsData, us_state_map, by="region")
```

```{r echo = FALSE}
# Plot for Catastrophic tier 
p <- ggplot()+ geom_polygon(data=mapdata, aes(x= long, y=lat, group = group,
                                             fill = mapdata$CatasTierCov), colour="white") 
p1 <- p+geom_text(data=statename, aes(x=long, y=lat, label=region),  na.rm = T, size=1)+
  coord_map() + scale_fill_continuous(name = "Actuarial Value in %", 
          guide = guide_legend()) +
  theme(legend.position = "bottom", axis.text = element_blank()) +
  ggtitle("US Health Insurance Plan\nCatastrophic Coverage Tier Based on Actuarial Value")
print(p1)
ggsave("inscatascov.png", width = 5, height = 5)

```

```{r echo = FALSE}
# Plot for Bronze Coverage Tier
p <- ggplot()+ geom_polygon(data=mapdata, aes(x= long, y=lat, group = group,
                                              fill = mapdata$BronzeTierCov), colour="red") 
p1 <- p+geom_text(data=statename, aes(x=long, y=lat, label=region),  na.rm = T, size=1)+
  coord_map() + scale_fill_continuous(name = "Actuarial Value in %", 
                                      guide = guide_legend()) +
  theme(legend.position = "bottom", axis.text = element_blank()) +
  ggtitle("US Health Insurance Plan\nBronze Coverage Tier Based on Actuarial Value")
print(p1)
ggsave("insbronzecov.png", width = 5, height = 5)
```

```{r echo = FALSE}
# Plot for Silver Coverage Tier
p <- ggplot()+ geom_polygon(data=mapdata, aes(x= long, y=lat, group = group,
                                              fill = mapdata$SilverTierCov), colour="red") 
p1 <- p+geom_text(data=statename, aes(x=long, y=lat, label=region),  na.rm = T, size=1)+
  coord_map() + scale_fill_continuous(name = "Actuarial Value in %", 
                                      guide = guide_legend()) +
  theme(legend.position = "bottom", axis.text = element_blank()) +
  ggtitle("US Health Insurance Plan\nSilver Coverage Tier Based on Actuarial Value")

ggsave("inssilvercov.png", width = 5, height = 5)
print(p1)
```

```{r echo = FALSE}
# Plot for Gold Coverage Tier
p <- ggplot()+ geom_polygon(data=mapdata, aes(x= long, y=lat, group = group,
                                              fill = mapdata$GoldTierCov), colour="red") 
p1 <- p+geom_text(data=statename, aes(x=long, y=lat, label=region),  na.rm = T, size=1)+
  coord_map() + scale_fill_continuous(name = "Actuarial Value in %", 
                                      guide = guide_legend()) +
  theme(legend.position = "bottom", axis.text = element_blank()) +
  ggtitle("US Health Insurance Plan\nGold Coverage Tier Based on Actuarial Value")
print(p1)
ggsave("insgoldcov.png", width = 5, height = 5)

```

```{r echo = FALSE}
# Plot for Platinum Coverage Tier
p <- ggplot()+ geom_polygon(data=mapdata, aes(x= long, y=lat, group = group,
                                              fill = mapdata$PlatinumTierCov), colour="red") 
p1 <- p+geom_text(data=statename, aes(x=long, y=lat, label= region),  na.rm = T, size=1)+
  coord_map() + scale_fill_continuous(name = "Actuarial Value in %", 
                                      guide = guide_legend()) +
  theme(legend.position = "bottom", axis.text = element_blank()) +
  ggtitle("US Health Insurance Plan\nPlatinum Coverage Tier Based on Actuarial Value")
print(p1)
ggsave("insplatinumcov.png", width = 5, height = 5)
```



